---
title: "Enrichment and gene set analyses"
author: "Dane Grosvenor"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DEG to GSEA and Plots
## Setup
### Load Libraries

```{r load_libs, message=FALSE, warning=FALSE}

library(tidyverse)
library(limma)
library(SEARchways)
library(ComplexHeatmap)
library(circlize)

set.seed(101)

```

### Create output directories

```{r create_out_dirs, message=FALSE}

dir.create("figures/heatmaps", showWarnings = FALSE)
dir.create("figures/boxplots", showWarnings = FALSE)
dir.create("figures/boxplots/exp1", showWarnings = FALSE)
dir.create("figures/boxplots/exp2", showWarnings = FALSE)
dir.create("figures/bubbleplots", showWarnings = FALSE)
dir.create("results/enrichment_outputs", showWarnings = FALSE)

```

## Load Data

```{r load_data, message=FALSE}

# load RData w/ multiple objects
loadRData <- function(fileName) {
  load(fileName)
  get(ls()[ls() !="fileName"])
}

# load exp1.lme and exp2.lme
exp1.lme <- loadRData("results/lme_outputs/experiment1_lme.RData")
exp2.lme <- loadRData("results/lme_outputs/experiment2_lme.RData")

# load exp1 and exp2 voom objects
voom.exp1 <- loadRData("data_clean/voom_exp1.RData")
voom.exp2 <- loadRData("data_clean/voom_exp2.RData")

```

### Get Heat Map Cluster Function

```{r get_hm_clust_fx}

get_hm_clust <- function(dat, hm, dimension){

  cluster.result <- data.frame()

  # Rows of heatmap
  if(dimension == "row"){
    #Deal with single cluster results
    if(is.list(ComplexHeatmap::row_order(hm))){
      clust.tot <- length(ComplexHeatmap::row_order(hm))
    } else {
      clust.tot <- 1
    }

    for (i in 1:clust.tot){
      #Get row indices
      if(is.list(ComplexHeatmap::row_order(hm))){
        # Dane version of hm row order needs to be numeric (CHM v2.15.1)
        cluster.index <- ComplexHeatmap::row_order(hm)[[(i)]]
      } else {
        cluster.index <- ComplexHeatmap::row_order(hm)
      }
      
      #Clusters with >1 element
      if(length(cluster.index) > 1){
        #Pull clusters
        cluster.result <- t(t(row.names(dat[cluster.index,]))) %>%
          #Convert to data frame
          as.data.frame() %>%
          #row order within cluster
          mutate(row_within_cluster = 1:length(cluster.index)) %>%
          #add cluster name
          mutate(cluster = paste0("cluster", i)) %>%
          #Rename default column
          rename(row=V1) %>%
          #concatenate results
          bind_rows(cluster.result)
      } else {
        #Clusters with only 1 element
        cluster.result <- data.frame(row = rownames(dat)[cluster.index],
                                     row_within_cluster = 1,
                                     cluster = paste0("cluster", i)) %>%
          bind_rows(cluster.result)
      }
    }
  } else if(dimension == "col"){
    # Columns of heatmap

    #Deal with single cluster results
    if(is.list(column_order(hm))){
      clust.tot <- length(column_order(hm))
    } else {
      clust.tot <- 1
    }
    
    for (i in 1:clust.tot){
      #Get column indices
      if(is.list(column_order(hm))){
        cluster.index <- column_order(hm)[[(i)]]
      } else {
        cluster.index <- column_order(hm)
      }

      #Clusters with >1 element
      if(length(cluster.index) > 1){
        cluster.result <- t(t(colnames(dat[,cluster.index]))) %>%
          as.data.frame() %>%
          mutate(col_within_cluster = 1:length(cluster.index)) %>%
          mutate(cluster = paste0("cluster", i)) %>%
          rename(col=V1) %>%
          bind_rows(cluster.result)
      } else {
        #Clusters with only 1 element
        cluster.result <- data.frame(col = colnames(dat)[cluster.index],
                                     col_within_cluster = 1,
                                     cluster = paste0("cluster", i)) %>%
          bind_rows(cluster.result)
      }
    }
  } else{ stop("dimension must be one of row or col.") }

  cluster.result <- cluster.result[order(cluster.result$cluster),]

  return(cluster.result)
}

```

## Heatmaps with ComplexHeatMap

### Experiment 1 Heatmap

```{r exp1_hm, message=FALSE}

# specificaly look at 2hr WT to KO contrast & 6hr WT to KO contrast
# FDR < 0.25

exp1_contrast_filter_2hr <- exp1.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "2hr KO" & contrast_lvl == "2hr WT") %>%
  dplyr::pull(gene)

exp1_contrast_filter_6hr <- exp1.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "6hr KO" & contrast_lvl == "6hr WT") %>%
  dplyr::pull(gene)

exp1_contrast_filters <- c(exp1_contrast_filter_2hr, exp1_contrast_filter_6hr)

exp1_sig_genes <- exp1.lme$lme %>%
  dplyr::filter(variable == "Time:Genotype") %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(gene %in% exp1_contrast_filters) %>%
  dplyr::pull(gene)

#Format exp1 for heatmap
exp1.E <- as.data.frame(voom.exp1$E) %>% 
  tibble::rownames_to_column() %>% 
  #Rename libraries
  tidyr::pivot_longer(-rowname, names_to = "libID") %>% 
  dplyr::left_join(voom.exp1$targets %>% dplyr::select(libID, Genotype, Time, Animal)) %>% 
  dplyr::mutate(name = paste(Genotype, Time, Animal)) %>% 
  arrange(desc(Genotype), Time, Animal) %>% 
  dplyr::select(-c(libID, Genotype, Time, Animal)) %>% 
  tidyr::pivot_wider() %>% 
  #Filter to significant genes
  dplyr::filter(rowname %in% exp1_sig_genes) %>% 
  #Rename genes to symbols
  dplyr::left_join(voom.exp1$genes, by=c("rowname"="ensembl_gene_id")) %>% 
  tidyr::unnest(symbol) %>% 
  dplyr::select(-rowname) %>% 
  tibble::column_to_rownames("symbol") %>%
  as.matrix()

exp1.counts.scaled <- t(scale(t(exp1.E)))

# k = 6 clust dendrogram

exp1.row.dend <- exp1.counts.scaled %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::set("branches_k_color", k = 6)

# top bar annotation
anno_groups <- str_split(colnames(exp1.counts.scaled), pattern = " ")

ha = ComplexHeatmap::HeatmapAnnotation(
  Genotype = sapply(anno_groups,"[[",1),
  Time = sapply(anno_groups,"[[",2),
  col = list(Genotype = c("WT" = "#13D025", "KO" = "#4B0092"),
             Time = c("0hr" = "#994F00", "2hr" = "#006CD1", "6hr" = "#E1BE6A"))
)

experiment_1_heatmap <- ComplexHeatmap::Heatmap(
  exp1.counts.scaled,
  top_annotation = ha,
  cluster_columns = FALSE,
  cluster_rows = exp1.row.dend,
  row_split = 6,
  column_gap = unit(1, "mm"),
  row_gap = unit(2, "mm"),
  row_names_gp = grid::gpar(fontsize = 7),
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_names_rot = 90,
  heatmap_legend_param = list(title = "Log2 CPM")
  )

pdf(file = "figures/heatmaps/experiment_1_heatmap.pdf")
draw(experiment_1_heatmap)
dev.off()

```

### Experiment 2 Heatmap

```{r exp2_hm, message=FALSE}

# specificaly look at 24hr WT to KO contrast
# FDR < 0.25

exp2_contrast_filter_24hr <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "24hr KO" & contrast_lvl == "24hr WT") %>%
  dplyr::pull(unique(gene))
  
exp2_sig_genes <- exp2.lme$lme %>%
  dplyr::filter(variable == "Time:Genotype") %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(gene %in% exp2_contrast_filter_24hr) %>%
  dplyr::pull(gene)

#Format exp1 for heatmap
exp2.E <- as.data.frame(voom.exp2$E) %>% 
  tibble::rownames_to_column() %>% 
  #Rename libraries
  tidyr::pivot_longer(-rowname, names_to = "libID") %>% 
  dplyr::left_join(voom.exp2$targets %>% dplyr::select(libID, Genotype, Time, Animal)) %>% 
  dplyr::mutate(name = paste(Genotype, Time, Animal)) %>% 
  arrange(desc(Genotype),Time,Animal) %>% 
  dplyr::select(-c(libID, Genotype, Time, Animal)) %>% 
  tidyr::pivot_wider() %>% 
  #Filter to significant genes
  dplyr::filter(rowname %in% exp2_sig_genes) %>% 
  #Rename genes to symbols
  dplyr::left_join(voom.exp2$genes, by=c("rowname"="ensembl_gene_id")) %>% 
  tidyr::unnest(symbol) %>% 
  dplyr::select(-rowname) %>% 
  tibble::column_to_rownames("symbol") %>%
  as.matrix()

exp2_counts_scaled <- t(scale(t(exp2.E)))

exp2_row_dend <- exp2_counts_scaled %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::set("branches_k_color", k = 7)

# top bar annotation
anno_groups2 <- str_split(colnames(exp2_counts_scaled), pattern = " ")

ha2 = ComplexHeatmap::HeatmapAnnotation(
  Genotype = sapply(anno_groups2,"[[",1),
  `Time/Condition` = sapply(anno_groups2,"[[",2),
  col = list(Genotype = c("WT" = "#13D025", "KO" = "#4B0092"),
             `Time/Condition` = c("0hr" = "#994F00", "24hr" = "#006CD1", "24hr+antiIL2" = "#E1BE6A"))
)

experiment_2_heatmap <- ComplexHeatmap::Heatmap(
  exp2_counts_scaled,
  top_annotation = ha2,
  cluster_columns = FALSE,
  cluster_rows = exp2_row_dend,
  row_split = 7,
  column_gap = unit(1, "mm"),
  row_gap = unit(4 , "mm"),
  row_names_gp = grid::gpar(fontsize = 7),
  use_raster = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_names_rot = 90,
  heatmap_legend_param = list(title = "Log2 CPM"))

pdf(file = "figures/heatmaps/experiment_2_heatmap.pdf")
ComplexHeatmap::draw(experiment_2_heatmap)
dev.off()

```

### Get Gene Lists per Cluster

```{r get_hm_cl_exp1, message=FALSE}

get_hm_cl_exp1 <- get_hm_clust(dat = exp1.counts.scaled, 
               hm = ComplexHeatmap::draw(experiment_1_heatmap), 
               dimension = "row")

write_csv(get_hm_cl_exp1, file = "figures/heatmaps/exp1_genes_per_cluster.csv")

```

```{r get_hm_cl_exp2, message=FALSE}

get_hm_cl_exp2 <- get_hm_clust(dat = exp2_counts_scaled, 
               hm = ComplexHeatmap::draw(experiment_2_heatmap), 
               dimension = "row")

write_csv(get_hm_cl_exp2, file = "figures/heatmaps/exp2_genes_per_cluster.csv")

```

## Enrichment

### Experiment 1 Enrichment (GO and Hallmark)

```{r include=FALSE}
load("results/enrichment_outputs/all_enrich.RData")
```

```{r enr_exp1, message=FALSE, eval=FALSE}
exp1_GO <- SEARchways::BIGprofiler(gene_list = list(int = get_hm_cl_exp1$row),
                        ID = "SYMBOL",
                        species = "mouse",
                        category = "C5",
                        subcategory = "GO:BP")

exp1_Hallmark <- SEARchways::BIGprofiler(gene_list = list(int = get_hm_cl_exp1$row),
                        ID = "SYMBOL",
                        species = "mouse",
                        category = "H")

exp1_GO %>% 
  mutate(genes=as.character(genes)) %>% 
  write_csv(file = "results/enrichment_outputs/exp1_GO_enr.csv")
exp1_Hallmark %>% 
  mutate(genes=as.character(genes)) %>% 
  write_csv(file = "results/enrichment_outputs/exp1_Hallmark_enr.csv")

```

### Experiment 2 Enrichment (GO and Hallmark)

```{r enr_exp2, message=FALSE, eval=FALSE}

exp2_GO <- SEARchways::BIGprofiler(gene_list = list(int = get_hm_cl_exp2$row),
                        ID = "SYMBOL",
                        species = "mouse",
                        category = "C5",
                        subcategory = "GO:BP")

exp2_Hallmark <- SEARchways::BIGprofiler(gene_list = list(int = get_hm_cl_exp2$row),
                        ID = "SYMBOL",
                        species = "mouse",
                        category = "H")

exp2_GO %>% 
  dplyr::mutate(genes=as.character(genes)) %>% 
  write_csv(file = "results/enrichment_outputs/exp2_GO_enr.csv")

exp2_Hallmark %>% 
  dplyr::mutate(genes=as.character(genes)) %>% 
  write_csv(file = "results/enrichment_outputs/exp2_Hallmark_enr.csv")

```

```{r eval=FALSE}
save(exp1_GO, exp1_Hallmark, exp2_GO, exp2_Hallmark,
     file="results/enrichment_outputs/all_enrich.RData")
```

### Enrichment Bubble Plot

#### Hallmark Experiment 2

```{r exp2_H_bubbleplot, message=FALSE}

exp2_H_enr_df <- exp2_Hallmark %>%
  dplyr::mutate(pathway = gsub('HALLMARK_', '', pathway)) %>%
  dplyr::mutate(pathway = gsub('_', ' ', pathway)) %>%
  dplyr::filter(group_in_pathway > 5 & FDR < 0.5)

### reference FC is WT minus KO, FC title change positive = down, negative = up

# 0hr DOWN in KO
bp_0hr_dn <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "0hr KO") %>%
  dplyr::filter(contrast_lvl == "0hr WT") %>%
  dplyr::filter(estimate > 0) %>%
  dplyr::pull(unique(gene))

# 0hr UP in KO
bp_0hr_up <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "0hr KO") %>%
  dplyr::filter(contrast_lvl == "0hr WT") %>%
  dplyr::filter(estimate < 0) %>%
  dplyr::pull(unique(gene))

### 24hr KO v 24hr WT
# 24hr UP in KO
bp_24hr_dn <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "24hr KO") %>%
  dplyr::filter(contrast_lvl == "24hr WT") %>%
  dplyr::filter(estimate > 0) %>%
  dplyr::pull(unique(gene))
       
# 24hr UP in KO
bp_24hr_up <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "24hr KO") %>%
  dplyr::filter(contrast_lvl == "24hr WT") %>%
  dplyr::filter(estimate < 0) %>%
  dplyr::pull(unique(gene))

### 24hr+anti-IL2 KO v 24hr+anti-IL2 WT
# 24hr+anti-IL2 UP in KO
bp_24hrIL2_up <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "(24hr+antiIL2 KO)") %>%
  dplyr::filter(contrast_lvl == "(24hr+antiIL2 WT)") %>%
  dplyr::filter(estimate < 0) %>%
  dplyr::pull(unique(gene))

# 24hr+anti-IL2 DOWN in KO
bp_24hrIL2_dn <- exp2.lme$lme.contrast %>%
  dplyr::filter(FDR < 0.25) %>%
  dplyr::filter(contrast_ref == "(24hr+antiIL2 KO)") %>%
  dplyr::filter(contrast_lvl == "(24hr+antiIL2 WT)") %>%
  dplyr::filter(estimate > 0) %>%
  dplyr::pull(unique(gene))

#Run enrichment
enr_all <- SEARchways::BIGprofiler(
  gene_list = list("0hr\nDOWN" = bp_0hr_dn,
                   "0hr\nUP" = bp_0hr_up,
                   "24hr\nDOWN" = bp_24hr_dn,
                   "24hr\nUP" = bp_24hr_up,
                   "24hr+anti-IL2\nDOWN" = bp_24hrIL2_dn,
                   "24hr+anti-IL2\nUP" = bp_24hrIL2_up),
  ID = "ENSEMBL",
  species = "mouse",
  category = "H")

signif_pw <- enr_all %>% 
  #24 hr contrasts only
  dplyr::filter(group %in% c("24hr\nDOWN", "24hr\nUP", 
                      "24hr+anti-IL2\nDOWN", "24hr+anti-IL2\nUP")) %>%
  dplyr::filter(group_in_pathway > 5) %>%
  dplyr::filter(FDR < 0.25) %>%
  pull(pathway) %>% unique()

exp2_enr_bp_df <- enr_all %>%
  dplyr::filter(pathway %in% signif_pw)  %>% 
  #24 hr contrasts only
  dplyr::filter(group %in% c("24hr\nDOWN", "24hr\nUP", 
                      "24hr+anti-IL2\nDOWN", "24hr+anti-IL2\nUP")) %>%
  dplyr::mutate(pathway = gsub("HALLMARK_", "", pathway)) %>%
  dplyr::mutate(pathway = gsub("_", " ", pathway)) %>% 
  dplyr::mutate(contrast = factor(group, 
                         levels = c("24hr\nDOWN", "24hr\nUP", 
                                      "24hr+anti-IL2\nDOWN",
                                    "24hr+anti-IL2\nUP"))) %>% 
  #signif group
  mutate(`FDR < 0.25` = ifelse(FDR<0.25,"Significant","Non-significant"))

# negative log 10 for scale

### Bubble Plot

exp2_bubble_plot <- ggplot(exp2_enr_bp_df, aes(y = pathway, x = contrast, size = group_in_pathway, shape=`FDR < 0.25`)) +
  geom_point(aes(color = -log10(FDR)), alpha = 1.0) +
  # theme(axis.text.x = element_text(angle = 90)) +
  labs(size = "Number of genes in pathway") +
  scale_x_discrete(drop = FALSE) +
  xlab("") +
  ylab("") +
  theme(panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  labs(size='# genes per pathway') +
  scale_shape_manual(values = c(21,16))

save(enr_all, file = "results/enrichment_outputs/exp2_enr_all.RData")

ggsave("figures/bubbleplots/exp2_enr_bubbleplot.pdf", exp2_bubble_plot,
       width = 8, height = 5)

exp2_bubble_plot

```

## Boxplots

### Experiment 1 all genes

```{r exp1_all_bps, message=FALSE, eval=FALSE}

exp1_gene_counts <- as.data.frame(exp1.E) %>%
  tibble::rownames_to_column("gene_symbol") %>%
  tidyr::pivot_longer(-gene_symbol) %>%
  dplyr::select(gene_symbol, name, value) %>%
  dplyr::arrange(name) %>%
  dplyr::mutate(new_name = (sub("^(\\S*\\s+\\S+).*", "\\1", name)),
                new_name = factor(new_name, 
                                  levels=c("WT 0hr","WT 2hr","WT 6hr",
                                           "KO 0hr","KO 2hr","KO 6hr")))

uniq_gene_exp1 <- unique(exp1_gene_counts$gene_symbol)

for (gene in seq_along(uniq_gene_exp1)) {
  
  gene_title <- uniq_gene_exp1[gene]
  # print(gene)
  
  temp_clust_num <- get_hm_cl_exp1  %>%
    dplyr::filter(row == gene_title) %>%
    dplyr::select(cluster) %>%
    dplyr::pull(cluster)
  
  ggplot(exp1_gene_counts %>% 
           dplyr::filter(gene_symbol == uniq_gene_exp1[gene]), 
         aes(x = new_name, y = value)) +
    geom_boxplot() +
    geom_point(aes(col = new_name)) +
    theme_bw() +
    theme(axis.text = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    ggtitle(paste0("Experiment 1 ", gene_title, " ", temp_clust_num)) +
    xlab("Genotype Time") +
    ylab(paste0(gene_title, " expression (log2)")) 
  
  filename <- paste0("figures/boxplots/exp1/exp1_", uniq_gene_exp1[gene], "_", temp_clust_num, ".png")
  
  ggsave(filename = filename, width = 4, height = 4)

}
beepr::beep()
```

### Experiment 2 all genes

```{r exp2_all_bps, message=FALSE, eval=FALSE}

exp2_gene_counts <- as.data.frame(exp2.E) %>%
  tibble::rownames_to_column("gene_symbol") %>%
  tidyr::pivot_longer(-gene_symbol) %>%
  dplyr::select(gene_symbol, name, value) %>%
  dplyr::arrange(name) %>%
  dplyr::mutate(new_name = (sub("^(\\S*\\s+\\S+).*", "\\1", name)))

uniq_gene_exp2 <- unique(exp2_gene_counts$gene_symbol)

# length(uniq_gene_exp2) # 1553

for (gene in seq_along(uniq_gene_exp2)) {
  
  gene_title <- uniq_gene_exp2[gene]
  
  temp_clust_num <- get_hm_cl_exp2  %>%
    dplyr::filter(row == gene_title) %>%
    dplyr::select(cluster) %>%
    dplyr::pull(cluster)
  
  ggplot(exp2_gene_counts %>% 
           filter(gene_symbol == uniq_gene_exp2[gene]), 
         aes(x = new_name, y = value)) +
    geom_boxplot() +
    geom_point(aes(col = new_name)) +
    theme(axis.text = element_text(angle = 90, vjust = 0.5)) +
    ggtitle(paste0("Experiment 2 ", gene_title, " ", temp_clust_num)) +
    xlab("Genotype:Time/Treatment") +
    ylab(paste0(gene_title, " expression (log2)"))
  
  filename <- paste0("figures/boxplots/exp2/exp2_", uniq_gene_exp2[gene], "_", temp_clust_num, ".png")
  
  ggsave(filename = filename, width = 7, height = 7)

}

```

```{r sesh_info}

sessionInfo()

```

